---
title: "Dataset Version Tracking with manifest.yaml"
author: "John Ehrlinger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset Version Tracking with manifest.yaml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## The Problem: Datasets Drift

A dataset pulled from Epic today is **not** the same file you will get if you
pull it again tomorrow — even when no one intentionally changed anything.
Backfilled lab results, corrected ICD codes, and updated patient status all
mean that re-running an extraction silently changes the underlying numbers.

The standard defence is a three-step protocol:

1. **Name every dataset by its extract date** — `cohort_20240115.csv`,
   `labs_20240801.sas7bdat`.  Never overwrite an existing dated file; a new
   pull gets a new filename.
2. **Record a SHA-256 checksum** the moment a file lands on disk.  A single
   changed byte produces a completely different hash.
3. **Verify the checksums before every analysis run** so that any drift is
   caught immediately, before results are generated.

`hvtiRutilities` provides two functions that implement this protocol:

| Function | Purpose |
|---|---|
| `update_manifest()` | Register a dataset (or update its entry) in `manifest.yaml` |
| `verify_manifest()` | Check every registered dataset before an analysis runs |

## Setup

```{r}
if (requireNamespace("hvtiRutilities", quietly = TRUE)) {
  library("hvtiRutilities")
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

For this vignette we work in a temporary directory that mimics a real project's
`datasets/` folder.

```{r}
datasets_dir <- file.path(tempdir(), "datasets")
dir.create(datasets_dir, showWarnings = FALSE)
manifest_path <- file.path(tempdir(), "manifest.yaml")
```

## Registering Datasets

### CSV — automatic row count

Generate a synthetic cohort (using `generate_survival_data()`) and write it as
a date-stamped CSV.

```{r}
set.seed(42)
cohort <- generate_survival_data(n = 200, seed = 42)

cohort_file <- file.path(datasets_dir, "cohort_20240115.csv")
write.csv(cohort, cohort_file, row.names = FALSE)
```

Register it in the manifest.  Row count is detected automatically from the CSV
header; only the SHA-256 is computed from the raw bytes.

```{r}
update_manifest(
  file         = cohort_file,
  manifest_path = manifest_path,
  extract_date = "2024-01-15",
  source       = "Epic EMR, query v4.2, ICD mapping v3.2",
  sort_key     = "ccfid"
)
```

Add a second file — a simulated labs extract pulled on the same date.

```{r}
set.seed(7)
labs <- data.frame(
  ccfid   = cohort$ccfid,
  hgb     = round(rnorm(200, mean = 13.5, sd = 1.8), 1),
  egfr    = round(rnorm(200, mean = 72,   sd = 18),   0),
  bnp     = round(exp(rnorm(200, mean = 4.8, sd = 0.9)), 0)
)

labs_file <- file.path(datasets_dir, "labs_20240115.csv")
write.csv(labs, labs_file, row.names = FALSE)

update_manifest(
  file          = labs_file,
  manifest_path = manifest_path,
  extract_date  = "2024-01-15",
  source        = "Epic EMR, labs module — Hgb, eGFR, BNP",
  sort_key      = "ccfid"
)
```

### Inspecting the manifest

The manifest is a plain YAML file committed to version control alongside your
analysis scripts (but **not** alongside the data files, which stay out of git).

```{r}
cat(paste(readLines(manifest_path), collapse = "\n"))
```

Each entry records exactly what is needed to re-verify the file later:
`extract_date`, `n_rows`, `sha256`, and any provenance notes you supply.

### SAS datasets — automatic row count

Files exported directly from SAS (`.sas7bdat`) are handled the same way.
`update_manifest()` calls `haven::read_sas()` internally to count rows;
the checksum is always computed from the raw file bytes.

```{r eval=FALSE}
update_manifest(
  file          = here::here("datasets", "cohort_20240115.sas7bdat"),
  manifest_path = here::here("manifest.yaml"),
  extract_date  = "2024-01-15",
  source        = "SAS CORR registry extract, query v3.1",
  sort_key      = "pat_id"
)
```

### Excel workbooks — automatic row count

Adjudication logs, crosswalk tables, and other curator-maintained datasets
often arrive as `.xlsx` files.  `update_manifest()` calls
`readxl::read_excel()` to count rows.

```{r eval=FALSE}
update_manifest(
  file          = here::here("datasets", "adjudication_20240115.xlsx"),
  manifest_path = here::here("manifest.yaml"),
  extract_date  = "2024-01-15",
  source        = "Clinical Events Committee adjudication log, v2"
)
```

### Other formats — supply `n_rows` explicitly

For RDS, Feather, Parquet, or any format not listed above, pass `n_rows`
directly.  The SHA-256 is still computed from the raw bytes.

```{r eval=FALSE}
update_manifest(
  file          = here::here("datasets", "cohort_20240115.rds"),
  manifest_path = here::here("manifest.yaml"),
  extract_date  = "2024-01-15",
  n_rows        = nrow(readRDS(here::here("datasets", "cohort_20240115.rds"))),
  source        = "RDS cached from Epic pull"
)
```

## Verifying Before Every Analysis

Place `verify_manifest()` at the very top of every analysis script and Quarto
document — before any data are loaded and before any results are produced.

```{r}
verify_manifest(
  manifest_path = manifest_path,
  data_dir      = datasets_dir
)
```

The function prints one confirmation line per file and returns invisibly.  If
everything is clean, your script continues.

### Typical script header

```{r eval=FALSE}
library(hvtiRutilities)
library(here)

# Abort immediately if any registered dataset has changed
verify_manifest(here("manifest.yaml"))

# Only reached if all checksums pass
cohort <- read.csv(here("datasets", "cohort_20240115.csv"))
labs   <- read.csv(here("datasets", "labs_20240115.csv"))
```

## What Happens When Data Changes

Suppose the data warehouse silently updates `cohort_20240115.csv` overnight.
`verify_manifest()` catches it the moment the next analysis run starts.

```{r error=TRUE}
# Simulate a silent dataset change
cohort_modified <- cohort
cohort_modified$age[1] <- cohort_modified$age[1] + 1L
write.csv(cohort_modified, cohort_file, row.names = FALSE)

# verify_manifest stops immediately
verify_manifest(
  manifest_path = manifest_path,
  data_dir      = datasets_dir
)
```

The error message names the affected file and shows both the expected and
actual SHA-256 values, giving you an unambiguous audit trail.

### Collecting all failures at once

During development it can be useful to see every problem before fixing any of
them.  Set `stop_on_error = FALSE` to collect warnings instead of stopping.

```{r warning=TRUE}
report <- verify_manifest(
  manifest_path = manifest_path,
  data_dir      = datasets_dir,
  stop_on_error = FALSE
)

report[report$status == "FAIL", c("file", "message")]
```

### Updating the manifest after a legitimate re-pull

If a new extract is intentional — a corrected pull, an updated query — call
`update_manifest()` again with the same filename.  The existing entry is
replaced in place and the new checksum is recorded.

```{r}
# Re-register after the legitimate change
update_manifest(
  file          = cohort_file,
  manifest_path = manifest_path,
  extract_date  = "2024-01-15",
  source        = "Epic EMR, query v4.2, ICD mapping v3.2 — corrected age for pt 1"
)

# Verification now passes again
verify_manifest(
  manifest_path = manifest_path,
  data_dir      = datasets_dir
)
```

## A New Pull for a Different Project

When the same registry is pulled again months later for a different analysis,
it gets a **new filename** — never overwriting the original.

```{r}
set.seed(99)
cohort_aug <- generate_survival_data(n = 315, seed = 99)

cohort_aug_file <- file.path(datasets_dir, "cohort_20240801.csv")
write.csv(cohort_aug, cohort_aug_file, row.names = FALSE)

update_manifest(
  file          = cohort_aug_file,
  manifest_path = manifest_path,
  extract_date  = "2024-08-01",
  source        = "Epic EMR, query v4.2, ICD mapping v3.3 — LVAD sub-study cohort",
  sort_key      = "ccfid"
)
```

The manifest now tracks all three files independently.

```{r}
m <- yaml::read_yaml(manifest_path)
do.call(rbind, lapply(m$datasets, function(e)
  data.frame(file = e$file, extract_date = e$extract_date,
             n_rows = e$n_rows, stringsAsFactors = FALSE)
))
```

Verification covers all of them in a single call.

```{r}
verify_manifest(
  manifest_path = manifest_path,
  data_dir      = datasets_dir
)
```

## Policy Recommendations

| Stage | Action |
|---|---|
| **On every data pull** | Run `update_manifest()` immediately and commit `manifest.yaml` to git |
| **Top of every analysis script** | Call `verify_manifest()` — before loading any data |
| **At manuscript submission** | Tag the git commit; freeze `manifest.yaml`; do not re-pull data |
| **For a new project or re-pull** | Use a new date-stamped filename; add a new `update_manifest()` entry |
| **`.gitignore`** | Add `datasets/` — commit the manifest, not the data |

## Summary

- `update_manifest()` records the SHA-256 checksum, row count, extract date,
  and optional provenance fields for CSV, SAS, Excel, or any file type.
- `verify_manifest()` checks every registered dataset before results are
  produced; one changed byte triggers an immediate, informative error.
- The combination makes dataset drift **detectable** and the audit trail
  **reproducible** from first pull to manuscript submission.

## Session Information

```{r}
sessionInfo()
```
