---
title: "Data Label Handling Best Practices"
author: "John Ehrlinger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Label Handling Best Practices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Why Labels Matter

Variable labels are the bridge between raw data and human-readable output.
A column named `lvefvs_b` means nothing in a table or plot caption, but
"Baseline LV Ejection Fraction (%)" tells the reader exactly what they are
looking at. In clinical research, where results go directly into
manuscripts, grant submissions, and regulatory filings, unlabeled output is
unprofessional and error-prone.

SAS datasets carry variable labels natively. When you import them into R
with `haven::read_sas()`, the `labelled` package preserves those labels as
column attributes. `hvtiRutilities` provides a set of functions to extract,
look up, register, and override these labels throughout the analysis
lifecycle.

```{r}
library(hvtiRutilities)
library(labelled)
```

## The Label Lifecycle

A typical clinical analysis has four phases where labels matter:

1. **Ingestion** --- labels arrive with the data (SAS import) or need to be
   created (CSV import)
2. **Transformation** --- derived variables (ratios, bins, indices) need new
   labels
3. **Override** --- study-specific abbreviations or corrections
4. **Consumption** --- labels appear in plots, tables, and data dictionaries

### Phase 1: Extracting Labels at Ingestion

`label_map()` extracts all variable labels into a lookup table:

```{r}
# Simulated SAS-style dataset with labels
dta <- generate_survival_data(n = 200, seed = 42)

lmap <- label_map(dta)
head(lmap, 10)
```

The result is a two-column data frame: `key` (variable name) and `label`
(descriptive text). Every column in the data gets an entry. Unlabeled
columns fall back to the variable name itself.

#### Detecting Missing Labels

When data arrives from a plain CSV (no labels), `label_map()` warns you:

```{r warning=TRUE}
csv_data <- data.frame(
  pat_id = 1:5,
  hgb = c(12.1, 14.3, 10.8, 13.5, 11.2),
  egfr = c(88, 72, 45, 91, 63)
)

# This triggers a warning --- most columns lack labels
lmap_csv <- label_map(csv_data)
print(lmap_csv)
```

The warning is intentional: it prevents the common mistake of generating
plots and tables with cryptic variable names like `hgb` and `egfr` and
not noticing until a collaborator asks what they mean.

### Phase 2: Labeling Derived Variables

After ingestion, you'll create new variables --- age groups, risk scores,
ratios, indicator flags. These variables have no labels because they didn't
exist in the original data.

#### Method A: Label the data directly (preferred)

The best practice is to label the data frame itself using `add_labels()`.
Labels travel with the data through `dplyr` operations, so they are never
out of sync:

```{r}
# Create derived variables
dta$age_group <- cut(dta$age,
  breaks = c(0, 30, 50, 70, Inf),
  labels = c("<30", "30-50", "50-70", ">70")
)
dta$ef_low <- dta$lvefvs_b < 40

# Label them directly on the data frame
dta <- add_labels(dta, c(
  age_group = "Age Group at Surgery",
  ef_low    = "Reduced Ejection Fraction (<40%)"
))

# Verify
var_label(dta$age_group)
var_label(dta$ef_low)
```

When you label the data directly, any subsequent call to `label_map()`
automatically picks up the new labels:

```{r}
lmap <- label_map(dta)
lmap[lmap$key %in% c("age_group", "ef_low"), ]
```

#### Method B: Update the label map (for reporting)

Sometimes the label map is created once and passed to a reporting module
that generates tables and figures. In that case, update the map directly:

```{r}
# Start from the base data
lmap <- label_map(generate_survival_data(n = 100, seed = 7))

# Register labels for variables you plan to create
lmap <- add_labels(lmap, c(
  age_group  = "Age Group at Surgery",
  ef_low     = "Reduced Ejection Fraction (<40%)",
  risk_score = "Composite Risk Score"
))

tail(lmap, 4)
```

#### Which method should I use?

| Scenario | Method | Reason |
|---|---|---|
| Adding columns to a data frame | `add_labels(data, ...)` | Labels travel with the data |
| Building a reporting table/dictionary | `add_labels(lmap, ...)` | The map is the artifact being consumed |
| Quick lookup for a plot title | `get_label(lmap, var)` | Safe, readable, errors on typos |

### Phase 3: Study-Specific Overrides

Different studies use different abbreviations and naming conventions. An
AVSD study might want "Common AVV" shortened to "CAVV"; a mitral valve
study might need different terminology entirely. Hard-coding these
replacements in shared functions is fragile.

`apply_label_overrides()` reads a YAML file and applies the overrides to
a label map. If the file doesn't exist, nothing happens --- making it safe
to call unconditionally.

```{r}
# Create a study-specific overrides file
tmp_overrides <- tempfile(fileext = ".yml")
writeLines(c(
  "lvefvs_b: 'Baseline LVEF (%)'",
  "hgb_bs: 'Hemoglobin (g/dL)'",
  "gfr_bs: 'eGFR (mL/min/1.73m2)'",
  "nyha_class: 'NYHA Class'"
), tmp_overrides)

# Start with the default labels
lmap <- label_map(generate_survival_data(n = 50, seed = 1))

# Apply study-specific overrides
lmap <- apply_label_overrides(lmap, overrides_file = tmp_overrides)
lmap[lmap$key %in% c("lvefvs_b", "hgb_bs", "gfr_bs", "nyha_class"), ]
```

```{r include=FALSE}
unlink(tmp_overrides)
```

In a real project, `labels_overrides.yml` lives alongside `config.yml` in
the project root and is committed to version control. Each study gets its
own file; shared analysis code never contains hard-coded label replacements.

#### Example `labels_overrides.yml`

```yaml
# AVSD study label overrides
cavv_area: "Common AVV Area (cm2)"
ed_area: "End-Diastolic Area (cm2)"
es_area: "End-Systolic Area (cm2)"
bsa_idx: "BSA-Indexed Value"
```

### Phase 4: Using Labels in Output

#### Safe single-variable lookup with `get_label()`

The `get_label()` function replaces the error-prone `match()` pattern. It
errors on typos instead of silently returning `NA`:

```{r}
lmap <- label_map(generate_survival_data(n = 50, seed = 42))

# Use in plot titles
get_label(lmap, "age")
get_label(lmap, "lvefvs_b")
```

```{r error=TRUE}
# Typo --- clear error instead of silent NA
get_label(lmap, "ages")
```

#### Building a data dictionary

```{r}
dta <- generate_survival_data(n = 200, seed = 42)
lmap <- label_map(dta)

# Build a type-annotated data dictionary
dict <- data.frame(
  variable    = lmap$key,
  label       = lmap$label,
  type        = vapply(dta, function(x) class(x)[1], character(1)),
  n_unique    = vapply(dta, function(x) length(unique(x)), integer(1)),
  pct_missing = vapply(dta, function(x) round(mean(is.na(x)) * 100, 1), numeric(1))
)
head(dict, 12)
```

#### Labels in summary tables

```{r}
# Numeric summary with labels
num_vars <- c("age", "bmi", "hgb_bs", "gfr_bs", "lvefvs_b")
summary_tbl <- data.frame(
  variable = num_vars,
  label    = vapply(num_vars, function(v) get_label(lmap, v), character(1)),
  mean     = vapply(num_vars, function(v) round(mean(dta[[v]]), 1), numeric(1)),
  sd       = vapply(num_vars, function(v) round(sd(dta[[v]]), 1), numeric(1)),
  median   = vapply(num_vars, function(v) round(median(dta[[v]]), 1), numeric(1))
)
print(summary_tbl)
```

#### Labels in plots

```{r fig.height=4}
var <- "lvefvs_b"
hist(dta[[var]],
  main = get_label(lmap, var),
  xlab = get_label(lmap, var),
  col  = "steelblue",
  border = "white",
  breaks = 20
)
```

## Complete Workflow

Putting it all together --- a realistic analysis setup:

```{r}
# 1. Load data (simulated here; in practice: haven::read_sas())
dta <- generate_survival_data(n = 500, seed = 2024)

# 2. Convert types
dta_clean <- r_data_types(dta,
  factor_size = 5,
  skip_vars   = c("ccfid", "iv_dead", "iv_reop", "iv_opyrs")
)

# 3. Create derived variables with labels
dta_clean$age_group <- cut(dta_clean$age,
  breaks = c(0, 30, 50, 70, Inf),
  labels = c("<30", "30-50", "50-70", ">70")
)
dta_clean$ef_category <- cut(dta_clean$lvefvs_b,
  breaks = c(0, 35, 50, Inf),
  labels = c("Reduced", "Borderline", "Normal")
)
dta_clean <- add_labels(dta_clean, c(
  age_group   = "Age Group at Surgery",
  ef_category = "Ejection Fraction Category"
))

# 4. Extract label map for reporting
lmap <- label_map(dta_clean)

# 5. Verify: all variables have real labels
stopifnot(!any(lmap$key == lmap$label))

# 6. Summary using labels
cat("Variables:", nrow(lmap), "\n")
cat("All labeled:", !any(lmap$key == lmap$label), "\n")
head(lmap, 10)
```

## Labels and `r_data_types()`

Labels are preserved through type conversion. This is handled automatically
--- you don't need to do anything special:

```{r}
dta <- sample_data(n = 50)

# Before conversion
var_label(dta$char)

# After conversion
dta_clean <- r_data_types(dta, skip_vars = "id")
var_label(dta_clean$char)

# Labels survive the round-trip
lmap_before <- label_map(dta)
lmap_after  <- label_map(dta_clean)
identical(lmap_before, lmap_after)
```

## Anti-Patterns to Avoid

### 1. Hard-coded label replacements in shared code

```{r eval=FALSE}
# BAD: study-specific logic in shared helper
clean_labels <- function(labels) {
  labels$label <- gsub("Common AVV", "CAVV", labels$label)
  labels$label <- gsub("End-distole", "End-Diastolic", labels$label)
  labels
}
```

This breaks the moment anyone uses the code for a different study. Use
`apply_label_overrides()` with a per-study YAML file instead.

### 2. Creating derived variables without labels

```{r eval=FALSE}
# BAD: new column has no label
dta$risk_score <- dta$age * 0.1 + as.integer(dta$nyha_class) * 0.5

# GOOD: label it immediately
dta$risk_score <- dta$age * 0.1 + as.integer(dta$nyha_class) * 0.5
dta <- add_labels(dta, c(risk_score = "Composite Risk Score"))
```

### 3. Using `match()` without error checking

```{r eval=FALSE}
# BAD: typo returns NA silently
title <- lmap$label[match("ages", lmap$key)]  # NA, no warning

# GOOD: get_label() catches the typo
title <- get_label(lmap, "ages")  # Error: 'ages' not found in label map
```

### 4. Labeling the map instead of the data

```{r eval=FALSE}
# FRAGILE: map goes stale when you modify the data
dta$ratio <- dta$a / dta$b
lmap <- add_labels(lmap, c(ratio = "A/B Ratio"))
# ... 50 lines later, someone renames 'ratio' to 'ab_ratio'
# lmap still says "ratio" --- silent mismatch

# BETTER: label the data, extract the map later
dta$ratio <- dta$a / dta$b
dta <- add_labels(dta, c(ratio = "A/B Ratio"))
lmap <- label_map(dta)  # always in sync
```

## Function Reference

| Function | Purpose |
|---|---|
| `label_map(data)` | Extract all labels into a lookup table |
| `get_label(lmap, var)` | Look up one label with error checking |
| `add_labels(data, labels)` | Label a data frame or update a label map |
| `apply_label_overrides(lmap, file)` | Apply study-specific overrides from YAML |

## Session Information

```{r}
sessionInfo()
```
