---
title: "PROC CONTENTS and PROC MEANS in R"
author: "John Ehrlinger"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{PROC CONTENTS and PROC MEANS in R}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load}
library(hvtiRutilities)
```

## Two Procedures You Already Run

Open a new extract in SAS and you almost certainly run two things before
anything else. `PROC CONTENTS` to see what arrived: how many observations, how
many variables, what each one is called and how it is stored. Then `PROC MEANS`
to see whether the numbers are plausible: means, ranges, how much is missing.

Neither is a modelling step. They are the sanity check you do before you trust
the file enough to analyse it. `proc_contents()` and `proc_means()` are those
two habits, kept intact in R.

The point is not that R lacks a way to describe a data frame. It has several.
The point is that you already know what these two tables should look like, and
when the R version prints the same numbers in the same shape, you can tell at a
glance whether the extract is right. A different summary in a different layout
makes you re-learn what "normal" looks like.

We will use the simulated cardiac surgery cohort that ships with the package,
so every example here runs as written:

```{r data}
dta <- generate_survival_data(n = 200, seed = 42)
```

## What Arrived? `proc_contents()`

In SAS you would write:

```sas
proc contents data=cohort;
run;
```

In R:

```{r contents-basic}
pc <- proc_contents(dta)
pc$header
```

The header is the block SAS prints at the top: observation count, variable
count, and the dataset label if the file carries one. Our simulated cohort has
no dataset label, so it reports `NA`.

The variables table is the *Alphabetic List of Variables and Attributes*.
Like SAS, it sorts by name by default:

```{r contents-vars}
head(proc_contents(dta)$variables, 6)
```

Pass `order = "varnum"` for creation order, the equivalent of the SAS
`VARNUM` option:

```{r contents-varnum}
head(proc_contents(dta, order = "varnum")$variables, 8)
```

Note that `num` always reports creation position, whichever sort you asked
for. Sorting alphabetically does not renumber the variables, so you can read a
name off the alphabetical list and still know where it sits in the record.

### Reading the Extra Columns

Three columns are not in the SAS output, and they are the reason to run this
in R rather than reading a `PROC CONTENTS` listing:

- **`class`** is the R class, sitting next to the SAS `type`. You want both,
  for reasons covered in the next section.
- **`n_unique`** counts distinct non-missing values. A numeric column with two
  distinct values is a flag someone forgot to declare; a character column with
  199 distinct values out of 200 rows is an identifier, not a category.
- **`pct_missing`** is completeness, and it is often the most useful number on
  the page. In our cohort:

```{r contents-missing}
pcv <- proc_contents(dta)$variables
pcv[pcv$pct_missing > 0, c("variable", "label", "pct_missing")]
```

`iv_reop` is 83.5% missing, and that is correct rather than broken: it is
time-to-reoperation, which only exists for the patients who had one. But you
want to *see* that before you put the variable in a model, not discover it when
the model silently drops 167 rows.

### Where the SAS Original Has More

`proc_contents()` does not report `Len`, `Pos`, or `Informat`, and it omits the
created and modified timestamps. This is a deliberate omission, not an
oversight.

Reading a `.sas7bdat` through `haven` is lossy. Variable name, label,
`format.sas`, and creation order survive the trip. Storage `LENGTH`, position
within the observation, and informat do not. We could infer a length from the
data, but an inferred value would look authoritative and would disagree with
the source dataset every time its `LENGTH` statement differed from the default.
A missing column is easier to work around than a wrong one that looks right.

So if you need the storage layout of the original file, read it from SAS. For
everything you actually use `PROC CONTENTS` for day to day, it is here.

### `Type` Is Two-Valued, and That Hides Things

SAS has two variable types, `Num` and `Char`. R has many. The mapping is
therefore lossy in the honest direction: `character` and `factor` report
`Char`, and everything else, including `logical`, `Date`, and `POSIXct`,
reports `Num`.

A SAS date *is* a number carrying a date format, so this is faithful. But it
means `type` alone cannot tell you a date from a lab value. That is exactly why
`class` sits beside it:

```{r contents-types}
dates <- data.frame(
  visit = as.Date(c("2024-01-15", "2024-02-20")),
  flag  = c(TRUE, FALSE),
  score = c(2.5, 3.1)
)
proc_contents(dates)$variables[, c("variable", "type", "class")]
```

Read the two columns together. `type` tells you what SAS would call it;
`class` tells you what R will actually do with it.

One edge case worth knowing before you hit it: for a data frame with columns
but zero rows, a fully filtered cohort say, `pct_missing` is `NaN` rather than
`0`. The proportion of missing values among no values is undefined, and
reporting `0` would claim that nothing is missing. If you are formatting this
column for a report, handle `NaN` explicitly.

## Do the Numbers Look Right? `proc_means()`

The SAS default:

```sas
proc means data=cohort;
run;
```

gives you N, mean, standard deviation, minimum, and maximum over every numeric
variable. So does the R default:

```{r means-default}
proc_means(dta, vars = c("age", "bmi", "gfr_bs"))
```

Omit `vars` and every numeric column is analysed, matching SAS's behaviour when
you leave out the `VAR` statement. Ask for specific statistics and you get them
in the order you asked, the same way a `PROC MEANS` statistic list works:

```{r means-stats}
proc_means(dta, vars = c("age", "bmi"),
           stats = c("n", "nmiss", "mean", "median", "q1", "q3"))
```

The keywords are SAS keywords: `n`, `nmiss`, `mean`, `std`, `min`, `max`,
`sum`, `range`, `stderr`, `cv`, `median`, `q1`, `q3`, and any `pNN` for a
percentile between 1 and 99.

```{r means-pctl}
proc_means(dta, vars = "bypass_time", stats = c("p5", "p50", "p95"))
```

### Stratifying with `class`

The `CLASS` statement becomes the `class` argument:

```sas
proc means data=cohort;
  class nyha_class;
  var age;
run;
```

```{r means-class}
proc_means(dta, vars = "age", class = "nyha_class",
           stats = c("n", "mean", "std", "median", "q1", "q3"))
```

Two behaviours here are inherited from SAS on purpose.

Rows with a missing value in any class variable are dropped, which is what
`PROC MEANS` does unless you add `MISSING`. And the class levels come out in
declared order rather than alphabetically, matching SAS's default
`ORDER=INTERNAL`. That is why NYHA class reads I, II, III, IV above instead of
sorting to I, II, III, IV by luck. Give the function an ordered factor and your
clinical scales stay in clinical sequence.

Rows are ordered by analysis variable first, then by class level, so a
multi-variable request stays readable:

```{r means-class-multi}
proc_means(dta, vars = c("age", "bmi"), class = "sex",
           stats = c("n", "mean", "std"))
```

### The Quartile Trap

This is the one thing in this vignette that will silently change your numbers,
so it gets its own section.

SAS and R do not compute quantiles the same way. There are nine defensible
estimators for a sample quantile; SAS defaults to `QNTLDEF=5`, and R's
`quantile()` defaults to `type = 7`. They are different estimators, and they
disagree.

Watch what happens on four observations:

```{r quantile-trap}
x <- c(1, 2, 3, 4)

# SAS QNTLDEF=5, which is what proc_means() uses
quantile(x, c(0.25, 0.5, 0.75), type = 2, names = FALSE)

# R's default
quantile(x, c(0.25, 0.5, 0.75), type = 7, names = FALSE)
```

The first quartile is 1.5 in SAS and 1.75 under R's default. The median agrees.
That is precisely why the discrepancy hides: you check the median, it matches,
you conclude the port is fine, and the quartiles in your Table 1 are quietly
off.

`proc_means()` uses `type = 2` throughout, so its `q1`, `q3`, `median`, and any
`pNN` reproduce SAS. On our cohort the gap is small but real:

```{r quantile-real}
proc_means(dta, vars = "age", stats = c("q1", "median", "q3"))

quantile(dta$age, c(0.25, 0.5, 0.75), type = 7, names = FALSE)
```

35.85 against 35.875. Small enough to survive rounding in a manuscript table,
large enough to fail a `PROC COMPARE` against the SAS baseline. The divergence
grows on small and even-numbered samples, which is to say on subgroups.

If you compute a quantile yourself anywhere else in the analysis, pass
`type = 2` and you will stay consistent with both `proc_means()` and the SAS
original.

### Logical Columns

R logicals are not `is.numeric()`, so they are left out of the default variable
set. Name one in `vars` and it is coerced to 0/1, which makes `mean` a
proportion, as `PROC MEANS` would report for a 0/1 numeric flag:

```{r means-logical}
flags <- data.frame(complication = c(TRUE, FALSE, TRUE, TRUE, FALSE))
proc_means(flags, vars = "complication", stats = c("n", "mean", "sum"))
```

## Did the Extract Change? `compare_datasets()`

`PROC CONTENTS` on the old file, `PROC CONTENTS` on the new one, then read the
two listings side by side looking for what moved. `compare_datasets()` does the
reading for you:

```{r compare}
v1 <- generate_survival_data(n = 100, seed = 1)
v2 <- generate_survival_data(n = 120, seed = 2)

# Pretend the new extract gained a variable and lost one
v2$creatinine <- rnorm(120, 1.1, 0.3)
v2$reop <- NULL

compare_datasets(v1, v2)
```

It reports four kinds of drift: row count, columns added or dropped, class
changes on shared columns, and label changes on shared columns. The last one
matters more than it sounds. A label that changes from "Baseline eGFR" to
"eGFR at discharge" is the same column name measuring a different thing, and
nothing but the label will tell you.

The full result is a list, so you can act on any piece programmatically:

```{r compare-parts}
diff <- compare_datasets(v1, v2)
diff$cols_added
diff$cols_dropped
diff$type_changes
```

This pairs directly with the manifest system described in the
[Dataset Version Tracking](dataset-versioning.html) vignette. The manifest tells
you *that* a file changed, by checksum. `compare_datasets()` tells you *what*
changed inside it.

## Where the R Version Differs

A summary of every place these functions depart from the SAS original, so you
can check the list rather than discover an entry mid-analysis.

| Behaviour | SAS | `hvtiRutilities` |
|---|---|---|
| Quantile estimator | `QNTLDEF=5` | `type = 2`, the same estimator |
| Default `PROC MEANS` statistics | N, Mean, Std Dev, Min, Max | Same five |
| Variables with no `VAR` statement | All numeric | All numeric (logicals excluded unless named) |
| Missing class levels | Dropped unless `MISSING` | Dropped |
| Class level order | `ORDER=INTERNAL` | Factor level order |
| `PROC CONTENTS` sort | Alphabetic by default | `order = "alpha"` by default |
| `Len`, `Pos`, `Informat` | Reported | Omitted, not recoverable through `haven` |
| Created / modified timestamps | Reported | Omitted, not recoverable through `haven` |
| `Type` for dates | `Num` | `Num`, with the R class in `class` |
| `pct_missing` on a zero-row table | n/a | `NaN`, not `0` |

## Checklist

- [ ] Run `proc_contents()` on every new extract before analysing it, the same
      way you would run `PROC CONTENTS`.
- [ ] Read `pct_missing` before choosing model variables, and confirm that any
      high-missing column is missing *by design*.
- [ ] Read `type` and `class` together; `type` alone cannot distinguish a date
      from a number.
- [ ] Use `proc_means()` quartiles rather than a bare `quantile()` call, and
      pass `type = 2` anywhere else you compute one.
- [ ] Declare ordered clinical scales as ordered factors so `class` output
      keeps its clinical sequence.
- [ ] Run `compare_datasets()` on every re-pull, and check `label_changes`,
      not just the column counts.

## See Also

- `?proc_contents`, `?proc_means`, `?compare_datasets`
- [Dataset Version Tracking](dataset-versioning.html) for the manifest system
  that pairs with `compare_datasets()`
- [Data Label Handling Best Practices](data-labels.html) for the labels these
  functions report
- [Reproducible Seeds in R and SAS](reproducible-seeds.html) for the other half
  of keeping an R analysis reconcilable against SAS
