---
title: "Simulating Cardiac Surgery Survival Data"
author: "John Ehrlinger"
date: today
format: html
vignette: >
  %\VignetteIndexEntry{Simulating Cardiac Surgery Survival Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `generate_survival_data()` function creates a realistic synthetic cardiac surgery cohort suitable for testing and demonstrating survival analysis workflows. Survival times are drawn from a Weibull distribution with a linear predictor built from clinical variables (LVEF, age, hemoglobin, NYHA class, eGFR), and administrative censoring is applied at up to 15 years.

```{r}
if (requireNamespace("hvtiRutilities", quietly = TRUE)) {
  library("hvtiRutilities")
} else {
  pkgload::load_all(export_all = FALSE, helpers = FALSE, quiet = TRUE)
}
```

## Generating the Dataset

```{r}
set.seed(42)
dta <- generate_survival_data(n = 500, seed = 1024)

dim(dta)
names(dta)
```

The dataset contains 24 columns covering patient identifiers, calendar anchors, survival outcomes, reoperation, demographics, pre-operative labs, cardiac function, and surgical variables.

## Data Structure

```{r}
str(dta)
```

Key columns:

| Column | Description |
|---|---|
| `ccfid` | Patient identifier |
| `origin_year` | Calendar year corresponding to `iv_opyrs = 0` |
| `iv_opyrs` | Observation interval length (years) anchored at `origin_year` |
| `iv_dead` | Observed follow-up time (years) |
| `dead` | Event indicator (1 = death, 0 = censored) |
| `reop` | Reoperation indicator |
| `iv_reop` | Time to reoperation (years; `NA` if no reoperation) |

## Outcome Summary

```{r}
# Event rates
cat("Death rate:       ", round(mean(dta$dead), 3), "\n")
cat("Reoperation rate: ", round(mean(dta$reop), 3), "\n")

# Follow-up distribution
summary(dta$iv_dead)
```

```{r}
hist(
  dta$iv_dead,
  breaks = 30,
  main = "Distribution of Follow-up Time",
  xlab = "Follow-up (years)",
  col = "steelblue",
  border = "white"
)
```

## Integration with `r_data_types()` and `label_map()`

The dataset arrives with variable labels attached and several columns that benefit from type conversion. The `r_data_types()` function handles these automatically.

```{r}
# Convert types: keep IDs and continuous outcomes as-is
model_data <- r_data_types(
  dta,
  factor_size = 5,
  skip_vars = c("ccfid", "iv_dead", "iv_reop", "iv_opyrs")
)

str(model_data[, c("dead", "reop", "sex", "nyha_class", "diabetes", "hypertension")])
```

After conversion:

- `dead` and `reop` (2 unique values) → `logical`
- `sex`, `diabetes`, `hypertension` (character) → `factor`
- `nyha_class` was already an ordered `factor`

### Extracting Variable Labels

```{r}
lmap <- label_map(model_data)
print(lmap)
```

The label map is useful for annotating tables and plots with descriptive names.

## Preparing Data for Survival Analysis

```{r}
# Drop admin columns to match a typical model_data setup
model_data <- model_data[, !names(model_data) %in% c("ccfid", "reop", "iv_reop",
  "origin_year", "iv_opyrs")]

# Outcome variable breakdown
table(model_data$dead)
```

```{r}
# Kaplan-Meier style summary by NYHA class (base R)
nyha_levels <- levels(model_data$nyha_class)

median_fu <- sapply(nyha_levels, function(lvl) {
  sub <- model_data[model_data$nyha_class == lvl, ]
  median(sub$iv_dead)
})

event_rate <- sapply(nyha_levels, function(lvl) {
  sub <- model_data[model_data$nyha_class == lvl, ]
  mean(sub$dead)
})

nyha_summary <- data.frame(
  nyha_class   = nyha_levels,
  n            = table(model_data$nyha_class),
  median_fu_yr = round(median_fu, 2),
  death_rate   = round(event_rate, 3)
)

print(nyha_summary)
```

## Reproducibility

The `seed` argument ensures reproducible datasets for testing:

```{r}
dta_a <- generate_survival_data(n = 100, seed = 99)
dta_b <- generate_survival_data(n = 100, seed = 99)

identical(dta_a, dta_b)  # TRUE
```

Different seeds produce different datasets:

```{r}
dta_c <- generate_survival_data(n = 100, seed = 7)
identical(dta_a, dta_c)  # FALSE
```

## Session Information

```{r}
sessionInfo()
```
